The Astropy Project: Building an open-science project and status of the v2.0 core package

### 2018-03-01 main.tex
address review comments

The \astropy Project aims to provide an open-source and open-development core
package (\astropypkg) and an ecosystem of \emph{affiliated packages} that
support astronomical functionality in the \python programming language.
The \astropypkg core package is now a feature-rich library of sufficiently
general tools and classes that supports the development of more specialized
code. An example of such functionality is reading and writing FITS files: It would be
time consuming and impractical for multiple groups to implement the FITS
standard \citep{FITS} and maintain software for such a general-purpose need.
Another example of such a common task is in dealing with representations of and
transformations between astronomical coordinate systems.

The \astropy Project aims to develop and provide high-quality code and
documentation according to the best practices in software development.
The project makes use of different tools and web services to reach those
goals without central institutional oversight.
The first public release of the \astropypkg package is described in
\cite{astropy}. Since then, the \astropypkg package has been
used in hundreds of projects and the scope of the package has grown
considerably. At the same time, the scientific community
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropypkg core has
developed.
In this paper, we describe the current status of the \astropy community and the
\astropypkg core package and discuss goals for future development.

We start by describing the way the \astropy Project functions and is organized
in \sectionname~\ref{sec:org}. We then describe the main software efforts
developed by the \astropy Project itself: a core package called \astropypkg
(\sectionname~\ref{sec:core}) and several separate packages that help maintain
the infrastructure for testing and documentation
(\sectionname~\ref{sec:infrastructure}). We end with a short vision for
the future of \astropy and astronomical software in general in
\sectionname~\ref{sec:future}. The full paper, including the code
to produce the figures, is available in a \github
repository\footnote{\url{https://github.com/astropy/astropy-v2.0-paper}}

@@

A major part of the \astropy Project is the concept of ``Affiliated Packages.''
An affiliated package is an astronomy-related
\python package that is not part of the \astropypkg core package, but
has requested to be included as part of the \astropy Project's
community. These packages support the goals and vision of \astropy of
improving code re-use, interoperability, and embracing good coding
practices such as testing and thorough documentation.

@@

Affiliated packages are listed on the main \astropy website and advertised to
the community through \astropy mailing lists; a list of current affiliated
packages is included in \tablename~\ref{tab:registry}.
Becoming an affiliated package is a good way for new and existing packages to
gain exposure while promoting \astropy's high standard for code and
documentation quality.
This process of listing and promoting affiliated packages is one way in which
the \astropy Project tries to increase code re-use in the astronomical
community.

@@

The \astropy Project, as of the version 2.0 release, does not receive any direct
financial support for the development of \astropypkg.
Development of the software, all supporting materials, and community support are
provided by individuals who work on the \astropy Project in their own personal
time, by individuals or groups contributing to \astropy as part of a research
project, or contributions from institutions that allocate people to work on
\astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgments.

Different funding models have been proposed for support of \astropy
(e.g., \citealt{2016arXiv161003159M}), but a long-term plan
for sustainability has not yet been established.
The \astropy Project has the ability to accept financial contributions
from institutions or individuals through the NumFOCUS\footnote{NumFOCUS
is a 501(c)(3) nonprofit that supports and promotes world-class, innovative,
open source scientific computing.} organization. NumFOCUS has, to date, covered the direct costs incurred by the \astropy Project.

\subsection{Reuse of the scientific Python ecosystem}

The \astropy Project is built on a philosophy of building from the existing
Python scientific ecosystem wherever possible. Many of the enabling technologies
like \texttt{numpy}, \texttt{scipy}, \texttt{matplotlib}, or \texttt{cython},
are necessary as the core underpinnings of \astropy packages.  Often this means
using these packages as ``building blocks'' (e.g., the ubiquitous
use of \texttt{numpy} arrays throughout \astropypkg).  In other cases this means
wrappers around more general algorithms that make them more convenient for
astronomy use cases (e.g., the model-fitting in the \package{astropy.modeling}
subpackage discussed in \sectionname~\ref{sec:modeling}).  Sometimes these
simple wrappers evolve into more complex implementations that address
astronomically-relevant use cases the general tool does not support (e.g.,
\package{astropy.convolution}, see \sectionname~\ref{ssec:convolution}). As a
broad rule, the Project explicitly encourages re-use of code where possible.

However, the boundaries of when this re-use is called for often ambiguous.  Some
of the examples above only evolved after significant debate in the community
over whether these algorithms were sufficient.  Other times, even apparently
general functionality did not exist in the wider ecosystem that met the \astropy
community's needs, and hence the functionality had to be developed wholly from
the developer resources available in the community (e.g.,
\package{astropy.units}, or \package{astropy.table} when it began). While some
of this functionality is ``given back'' to the wider community (e.g., myriad bug
fixes for the enabling technologies listed above, and some of the testing and
documentation architecture), in general, careful consideration is taken of
whether the additional effort to make a piece of \astropy, and it is only
``extracted'' when it seems plausible that the ongoing effort is worth the
reward in the form of additional contributors or maintainers.

\section{Astropy Core Package Version 2.0}
\label{sec:core}
The \astropy Project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version 0.2 (previously described in \citealt{astropy}).  The \astropypkg
package
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen improved performance
since the release of the version 0.2 package.

@@

In order to check the accuracy of the coordinate transformations in
\package{astropy.coordinates}, we have created a set of benchmarks that we use
to compare transformations between a set of coordinate frames for a number of
packages\footnote{\url{http://www.astropy.org/coordinates-benchmark/summary.html}}.
Since no package can be guaranteed to implement all transformations to
arbitrary precision and some transformations are sometimes subject to
interpretation of standards (in particular in the case of Galactic coordinates),
we do not designate any of the existing packages as the ``ground truth'' but
instead compare each tool to all other tools. The benchmarks are thus useful
beyond the \astropy Project since they allow all of the tools to be compared to
all other tools. The tools included in the benchmark at the moment include the
\astropypkg core package, Kapteyn \citep{kapteyn}, NOVAS \citep{novas},
PALpy \citep{pal}, PyAST \citep[a wrapper for AST, described in][]{ast},
PyTPM\footnote{\url{https://github.com/phn/pytpm}}, PyEphem \citep{pyephem},
and pySLALIB \citep[a Python wrapper for SLALIB, described in][]{slalib}.

@@

The \package{astropy.table} subpackage provides functionality for
representing and manipulating heterogeneous data. In some respects,
this is similar to \package{numpy} record arrays \citep{numpy} or
\package{pandas} \texttt{DataFrame} objects \citep{pandas}, but with modifications for
astronomical data. Most notably, tables from \package{astropy.table}
allow for table or column metadata and can handle
vectors/arrays or arbitrary objects (with suitable column-like characteristics)
as table entries. The subpackage was described in
detail in \cite{astropy}.  Thus, in what follows, we only summarize
key new features or updates to \package{astropy.table} since the
previous \astropy paper. These are support for grouped table
operations, table concatenation, and using array-valued
\package{astropy} objects as table columns.

@@

Additionally, there is now interoperability between \texttt{Table}
and \package{pandas} \texttt{DataFrame} objects using the \texttt{Table.to_pandas()}
and \texttt{Table.from_pandas()} methods. \texttt{Table.to_pandas()} is,
however, limited to a subset of possible \texttt{Table} objects, because several
of the features outlined above cannot be represented as \package{pandas} data
frames. While these features might be contributed to or combined with
\package{pandas} at a later date, the architectural differences that
enable these areas are significant, so this would be a major undertaking that would 
require substantial investment from the core maintainers and communities of both 
packages.

@@

\astropy provides a package template --- as a separate \github repository,
\package{astropy/package-template}\footnote{\url{https://github.com/astropy/package-template/}}
--- that aims to simplify setting up packaging, testing, and
documentation builds for developers of affiliated packages or
\astropypkg-dependent packages.
Any \python package can make use of this ready-to-go package layout, setup,
installation, and \package{Sphinx} documentation build infrastructure that was
originally developed for the \astropypkg core package and affiliated packages
maintained by the \astropy Project.
The package template also provides a testing framework, template configurations
for continuous integration services, and \package{Cython} build support.

@@

The documentation for many \python packages, including
all the packages in the \astropy ecosystem, is written using the
\package{Sphinx} documentation build system.
\package{Sphinx} supports writing documentation using plain text files
that follow a markup language called \texttt{reStructuredText} (RST).
These files are then transformed into HTML, PDF, or \LaTeX{} documents
during the documentation build process.
For the \astropy Project, we have developed several \package{Sphinx} extensions
that facilitate automatically generating API documentation for large projects,
like the \astropypkg core package.
The main extension we have developed is
\package{sphinx-automodapi}\footnote{\url{http://sphinx-automodapi.readthedocs.io}},
which provides a convenient single RST command to generate a set of
documentation pages, listing all of the available classes, functions, and
attributes in a given \python module.

@@

Beyond the core code, the \astropy Project is also further developing the
\astropy-managed affiliated packages.
While these may not be integrated into the \astropypkg core package, these
projects provide code that is useful to the astronomical community and meet the
testing and documentation standards of \astropy.
Some of these new efforts include an initiative to develop tools for
spectroscopy \citep[\package{specutils}, \package{specreduc}, \package{specviz}]
{ape13}, integration of LSST software, and support for HEALPIX
projection. %\citep{astropy-healpix}
%\inlinecomment{BMS}{TODO: uncomment references of packages once they are available}

@@

The development of the \astropypkg package and cultivation of the \astropy
ecosystem are still maintaining significant growth while improving in stability,
breadth, and reliability.
As the \astropypkg core package becomes more mature, several subpackages have
reached stability with a rich set of features that help astronomers worldwide to
perform many daily tasks, such as planning observations, analyzing data or
simulation results, and writing publications.
The strong emphasis that the \astropy Project puts on reliability and high coding
standards helps users to trust the calculations performed with \astropypkg and
to publish reproducible results.
At the same time, the \astropy ecosystem and core package are both growing: new
functionality is still being contributed and new affiliated packages are being
developed to support more specialized needs.

The \astropy Project is also spreading awareness of best practices in
community-driven software development.
This is important as most practicing astronomers were not explicitly taught
computer science and software development, despite the fact that a substantial
fraction of many astronomers' workload today is related to software use
and development.
The \astropypkg package leads by example, showing all interested astronomers how
modern tools like \texttt{git} version control or CI testing can
increase the quality, accessibility, and discoverability of astronomical
software without overly complicating the development cycle.
Within \astropy, all submitted code is reviewed by at least one (but typically
more than one) member of the \astropy community. Reviewers provide feedback
to contributors, which helps to improve contributors' software development skills.
As a community, \astropy follows an explicit code of conduct \citep{ape8} and treats all
contributors and users with respect, provides a harassment-free environment, and
encourages and welcomes new contributions from all.
Thus, while the \astropy Project provides and develops software and tools
essential to modern astronomical research, it also helps to prepare the current
and next generation of researchers with the knowledge to adequately use,
develop, and contribute to those tools within a conscientious and welcoming
community.

>>>>

The \astropy project aims to provide an open-source and open development core
package (\astropypkg) and an ecosystem of \emph{affiliated packages} that
support astronomical functionality in the \python programming language.
``Open development'' describes a process where anybody with an internet connection can suggest changes to the source code and contribute their opinion when new features, bug fixes, or other code changes, governance, or any other aspect of the development process is discussed (see Sect.~\ref{sect:devmod} and \ref{sect:ape} of how this is organized in practice).
The \astropypkg core package is now a feature-rich library of sufficiently
general tools and classes that supports the development of more specialized
code. An example of such functionality is reading and writing FITS files: It would be
time consuming and impractical for multiple groups to implement the FITS
standard \citep{FITS} and maintain software for such a general-purpose need.
Another example of such a common task is in dealing with representations of and
transformations between astronomical coordinate systems.

The \astropy project aims to develop and provide high-quality code and
documentation according to the best practices in software development.
The project makes use of different tools and web services to reach those
goals without central institutional oversight.
The first public release of the \astropypkg package is described in
\cite{astropy}. Since then, the \astropypkg package has been
used in hundreds of projects and the scope of the package has grown
considerably. At the same time, the scientific community
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropypkg core has
developed.
In this paper, we describe the current status of the \astropy community and the
\astropypkg core package and discuss goals for future development.

We start by describing the way the \astropy project functions and is organized
in \sectionname~\ref{sec:org}. We then describe the main software efforts
developed by the \astropy project itself: a core package called \astropypkg
(\sectionname~\ref{sec:core}) and several separate packages that help maintain
the infrastructure for testing and documentation
(\sectionname~\ref{sec:infrastructure}). We end with a short vision for
the future of \astropy and astronomical software in general in
\sectionname~\ref{sec:future}. The full paper, including the code
to produce the figures, is available in a \github
repository\footnote{\url{https://github.com/astropy/astropy-v2.0-paper}}

@@

A major part of the \astropy project is the concept of ``Affiliated Packages.''
An affiliated package is an astronomy-related
\python package that is not part of the \astropypkg core package, but
has requested to be included as part of the \astropy project's
community. These packages support the goals and vision of \astropy of
improving code re-use, interoperability, and embracing good coding
practices such as testing and thorough documentation.

@@

Affiliated packages are listed on the main \astropy website and advertised to
the community through \astropy mailing lists; a list of current affiliated
packages is included in \tablename~\ref{tab:registry}.
Becoming an affiliated package is a good way for new and existing packages to
gain exposure while promoting \astropy's high standard for code and
documentation quality.
This process of listing and promoting affiliated packages is one way in which
the \astropy project tries to increase code re-use in the astronomical
community.

@@

The \astropy project, as of the version 2.0 release, does not receive any direct
financial support for the development of \astropypkg.
Development of the software, all supporting materials, and community support are
provided by individuals who work on the \astropy project in their own personal
time, by individuals or groups contributing to \astropy as part of a research
project, or contributions from institutions that allocate people to work on
\astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgments.

Different funding models have been proposed for support of \astropy
(e.g., \citealt{2016arXiv161003159M}), but a long-term plan
for sustainability has not yet been established.
The \astropy project has the ability to accept financial contributions
from institutions or individuals through the NumFOCUS\footnote{NumFOCUS
is a 501(c)(3) nonprofit that supports and promotes world-class, innovative,
open source scientific computing.} organization. NumFOCUS has, to date, covered the direct costs incurred by the \astropy project.

\section{Astropy Core Package Version 2.0}
\label{sec:core}
The \astropy project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version 0.2 (previously described in \citealt{astropy}).  The \astropypkg
package
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen improved performance
since the release of the version 0.2 package.

@@

In order to check the accuracy of the coordinate transformations in
\package{astropy.coordinates}, we have created a set of benchmarks that we use
to compare transformations between a set of coordinate frames for a number of
packages\footnote{\url{http://www.astropy.org/coordinates-benchmark/summary.html}}.
Since no package can be guaranteed to implement all transformations to
arbitrary precision and some transformations are sometimes subject to
interpretation of standards (in particular in the case of Galactic coordinates),
we do not designate any of the existing packages as the ``ground truth'' but
instead compare each tool to all other tools. The benchmarks are thus useful
beyond the \astropy project since they allow all of the tools to be compared to
all other tools. The tools included in the benchmark at the moment include the
\astropypkg core package, Kapteyn \citep{kapteyn}, NOVAS \citep{novas},
PALpy \citep{pal}, PyAST \citep[a wrapper for AST, described in][]{ast},
PyTPM\footnote{\url{https://github.com/phn/pytpm}}, PyEphem \citep{pyephem},
and pySLALIB \citep[a Python wrapper for SLALIB, described in][]{slalib}.

@@

The \package{astropy.table} subpackage provides functionality for
representing and manipulating heterogeneous data. In some respects,
this is similar to \package{numpy} record arrays \citep{numpy} or
\package{pandas} dataframes \citep{pandas} but with modifications for
astronomical data. Most notably, tables from \package{astropy.table}
allow for table or column metadata and can handle
vectors or arrays as table entries.
The subpackage was described in
detail in \cite{astropy}.  Thus, in what follows, we only summarize
key new features or updates to \package{astropy.table} since the
previous \astropy paper. These are support for grouped table
operations, table concatenation, and using array-valued
\package{astropy} objects as table columns.

@@

@@

\astropy provides a package template --- as a separate \github repository,
\package{astropy/package-template}\footnote{\url{https://github.com/astropy/package-template/}}
--- that aims to simplify setting up packaging, testing, and
documentation builds for developers of affiliated packages or
\astropypkg-dependent packages.
Any \python package can make use of this ready-to-go package layout, setup,
installation, and \package{Sphinx} documentation build infrastructure that was
originally developed for the \astropypkg core package and affiliated packages
maintained by the \astropy project.
The package template also provides a testing framework, template configurations
for continuous integration services, and \package{Cython} build support.

@@

The documentation for many \python packages, including
all the packages in the \astropy ecosystem, is written using the
\package{Sphinx} documentation build system.
\package{Sphinx} supports writing documentation using plain text files
that follow a markup language called \texttt{reStructuredText} (RST).
These files are then transformed into HTML, PDF, or \LaTeX{} documents
during the documentation build process.
For the \astropy project, we have developed several \package{Sphinx} extensions
that facilitate automatically generating API documentation for large projects,
like the \astropypkg core package.
The main extension we have developed is
\package{sphinx-automodapi}\footnote{\url{http://sphinx-automodapi.readthedocs.io}},
which provides a convenient single RST command to generate a set of
documentation pages, listing all of the available classes, functions, and
attributes in a given \python module.

@@

Beyond the core code, the \astropy project is also further developing the
\astropy-managed affiliated packages.
While these may not be integrated into the \astropypkg core package, these
projects provide code that is useful to the astronomical community and meet the
testing and documentation standards of \astropy.
Some of these new efforts include an initiative to develop tools for
spectroscopy \citep[\package{specutils}, \package{specreduc}, \package{specviz}]
{ape13}, integration of LSST software, and support for HEALPIX
projection. %\citep{astropy-healpix}
%\inlinecomment{BMS}{TODO: uncomment references of packages once they are available}

@@

The development of the \astropypkg package and cultivation of the \astropy
ecosystem are still maintaining significant growth while improving in stability,
breadth, and reliability.
As the \astropypkg core package becomes more mature, several subpackages have
reached stability with a rich set of features that help astronomers worldwide to
perform many daily tasks, such as planning observations, analyzing data or
simulation results, and writing publications.
The strong emphasis that the \astropy project puts on reliability and high coding
standards helps users to trust the calculations performed with \astropypkg and
to publish reproducible results.
At the same time, the \astropy ecosystem and core package are both growing: new
functionality is still being contributed and new affiliated packages are being
developed to support more specialized needs.

The \astropy project is also spreading awareness of best practices in
community-driven software development.
This is important as most practicing astronomers were not explicitly taught
computer science and software development, despite the fact that a substantial
fraction of many astronomers' workload today is related to software use
and development.
The \astropypkg package leads by example, showing all interested astronomers how
modern tools like \texttt{git} version control or CI testing can
increase the quality, accessibility, and discoverability of astronomical
software without overly complicating the development cycle.
Within \astropy, all submitted code is reviewed by at least one (but typically
more than one) member of the \astropy community. Reviewers provide feedback
to contributors, which helps to improve contributors' software development skills.
As a community, \astropy follows an explicit code of conduct \citep{ape8} and treats all
contributors and users with respect, provides a harassment-free environment, and
encourages and welcomes new contributions from all.
Thus, while the \astropy project provides and develops software and tools
essential to modern astronomical research, it also helps to prepare the current
and next generation of researchers with the knowledge to adequately use,
develop, and contribute to those tools within a conscientious and welcoming
community.
